---
title: "Practical Machine Learning Course Assignment"
author: "Jill A Brown"
date: "May 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Problem

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.  One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The participants included in the data set were asked to perform barbell lifts correctly and incorrectly in 5 different ways and the goal was to predict the manner in which they did the exercise using data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants.  

```{r libraries, echo=FALSE, message=FALSE}
library(dplyr)
library(knitr)
library(ggplot2)
library(GGally)
library(AppliedPredictiveModeling)
library(caret)
library(Hmisc)
library(rpart)
library(rpart.plot)
```

## Read Data
First, we need to read in the data and clean it.  There are a number of columns with undefined variables and these will be removed.
```{r readdata, echo=FALSE}
activitydata <- read.csv("C:/Users/jilla/Google Drive/datasciencecoursera/machinelearning/pml-training.csv", header=TRUE, na.strings=c('','NA','#DIV/0!'))
activitydata2 <- activitydata[ , colSums(is.na(activitydata)) == 0]
dim(activitydata2)
attach(activitydata2)
```

After the data are read in, we need to create a training and testing set.  Though there's a test set for the final part of the assignment, we still want to be able to get an out of sample error rate.  The current training set is 70% of the provided training set; the test set is 30%.
```{r traintest}
set.seed(1026)
train = createDataPartition(y=activitydata2$classe, p=0.7, list=FALSE)
trainact = activitydata2[train,]
testact = activitydata2[-train,]
trainact2 <- trainact[c(8:60)]
testact2 <- testact[c(8:60)]
dim(trainact2); dim(testact2)
attach(trainact2)
```

## Explore Data
We want to look at the variables and the general structure of the data.
```{r explore, echo=FALSE}
str(trainact2)
head(trainact2)
```


## Build Model
Two models will be built to try to predict the class of exercise that is being performed.  The models being tested are the partition and a random forest.
``` {r buildpart, echo=FALSE}
modelact <- train(classe ~ ., method="rpart", data=trainact2, na.action=na.omit)
print(modelact$finalModel)
plot(modelact$finalModel, uniform=TRUE, main="Classification Tree")
text(modelact$finalModel, use.n=TRUE, all=TRUE, cex=.8)
```
``` {r buildrt, echo=FALSE}
modelact2 <- train(classe ~ ., method="rf", data=trainact2)
print(modelact2$finalModel)
```



## Cross Validation and Out of Sample Error
In order to cross validate we try both our models on the test set that was set aside at the beginning of this exploration.  First we will analyze the partition model.
``` {r testrpart}
predictrpart <- predict(modelact, newdata=testact2, type="raw")
matrixrpart <- confusionMatrix(predictrpart, testact2$classe)
matrixrpart
```
This model demonstrates a low degree of accuracy.

Now we will analyze the random forest model.
``` {r testrt}
predictrt <- predict(modelact2, newdata=testact2, type="raw")
matrixrt <- confusionMatrix(predictrt, testact2$classe)
matrixrt
```
This model demonstrates a high degree of accuracy.  Based on these cross validation activities and the out of sample error (accuracy value), it is recommended that a random forest model be used.

``` {r imp, echo=FALSE}
# estimate variable importance
importance <- varImp(modelact2, scale=FALSE)
plot(importance, main = "Top 5 Variables", top = 5)
```
The above plot shows the most important variables by rank in the random forest prediction model.


## Final Test Data
Now the random forest model will be used to predict the final twenty values.
``` {r finaltest}
activitydatatest <- read.csv("C:/Users/jilla/Google Drive/datasciencecoursera/machinelearning/pml-testing.csv", header=TRUE, na.strings=c('','NA','#DIV/0!'))
activitydatatest2 <- activitydatatest[ , colSums(is.na(activitydatatest)) == 0]
predictfinal <- predict(modelact2, newdata=activitydatatest2, type="raw")
predictfinal

```


## Citation

This dataset and corresponding information can be found here:

Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4hikb2jJM
